---
title: "Correlation"
author: "Remus"
date: "2022/2/7"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Correlation

Focus on 1) correlations between response variable and predictor variable (checking linearity)
         2) correlations between predictor variables (multicollinearity)
         
## Loading data and packages

```{r Loading, message = FALSE}
load("cancer.RData")
load("cancer2.RData")
library(dplyr)
library(GGally) 
library(ggplot2)
library(tidyr)  

```

## Using Alex's cancer2 (removed missing or incorrect values)
```{r cancer2}
cancer1 <- cancer
susPoint <- which(cancer1$AvgHouseholdSize < 0.5)
cancer1$AvgHouseholdSize[susPoint] <- NA
cancer2 <- na.omit(cancer1)
```

## Linearity

First we plot the response variable (deathRate) against every predictor variable.

```{r Linearity, echo=FALSE}
cancer2 %>% gather(-deathRate,-Geography,-binnedInc, key="predictor", value="value") %>%
  ggplot(aes(x = value, y=deathRate)) + facet_wrap(~predictor, scales = "free") +
  geom_point(size=0.35)
```

From the plots, we can see that most variables have a linear trend. (Use PctEmployed16_Over as an example?)
AvgHouseholdsize, MedianAgeFemale, MedianAgeMale, PctBlack and medIncome do not show clear linearity.
In particular, medIncome shows a exponential distribution and PctBlack does not show a relationship at all.
Therefore, PctBlack and medIncome need transformations.

```{r medIncome,echo=FALSE}
ggplot(cancer2,aes(x=medIncome, y=deathRate))+
  geom_point(size=0.8)
```
```{r PctBlack,echo=FALSE}
ggplot(cancer2,aes(x=PctBlack, y=deathRate))+
  geom_point(size=0.8)
```

## Correlation between predictor variables

First we plot the correlation matrix.
Don't know how to shift the labels.
```{r Correlation matrix, echo=FALSE}
ggcorr(cancer2[,-c(1,4)], label=TRUE,size = 2)
```

We can also see the scatter plots between the predictor variables.
```{r Correlation scatter,echo=FALSE}
ggpairs(cancer2[,-c(1,4)],
        upper = list(continuous = wrap("cor", size = 3)),
        lower = list(continuous = wrap("points", alpha = 0.3, size=0.01)))
```
We have the following pairs/clusters of highly correlated(greater than 0.7) 
predictor variables.

1) medIncome with
  povertyPercent, PctEmployed_Over16, PctPrivateCoverage, PctEmpPrivCoverage, PctPublicCoverage
  
medIncome and povertyPercent are measuring very similar things.
Higher median income means people are likely to afford private healthcare, and
do not require public heathcare.

Note that medIncome is highly correlated with PctEmployed_Over16 (0.7) but
not with PctUnemployed_Over16 (-0.5)

2) povertyPercent with 
  PctUnemployed_Over16, PctEmployed_Over16, PctPrivateCoverage, PctEmpPrivCoverage, PctPublicCoverage

Similar to medIncome. But we can see povertyPercent is highly correlated to PctUnemployed_Over16,
while medIncome does not.
  
3) MedianAgeMale with
  MedianAgeFemale
  
4) PercentMarried with
  PctMarriedHouseholds

These two pairs are each measuring almost the exact same thing.
  
6) PctEmployed16_Over with 
   PctPrivateCoverage, PctEmpPrivCoverage, PctPublicCoverage

A higher employment means they probably get private healthcare from their companies.
Simikar reasons to medIncome.
   
7) PctPrivateCoverage with
  PctEmpPrivCoverage, PctPublicCoverage
  
Healthcare coverage variables are measuring similar thing,

