---
title: "Skewness"
output: pdf_document
---
```{r, echo=FALSE}
library(ggplot2)
library(GGally)
library(moments)
```

# Findings - Skew

It is often helpful to transform the predictor variables to reduce skewness and make the distribution more centralised. 
From our analysis, an approximate benchmark figure for acceptable skew is [-0.6,+0.6]. 
Doing so would allow for an accurate modelling without drastically transforming the data and losing initial porperties of the data. However, this might be subjecive to the modeller. 

8 out of 17 predictor variables are already within the acceptable range for skewness, 1 predictor variable, binnedIncome, is categorical and Geography is classified as "character" and is not used as an explanatory variable. 
Therefore, this leaves us with 7 predictor variables to transform. Most are fixed using log or square root/cube root transformations.

            Variable           Transformation
            
1        IncidenceRate                 log
2            medIncome                 log
3       povertyPercent           Cube Root
4     AvgHouseholdSize                 log
5       PercentMarried Raise to power of 2
6 PctUnemployed16_Over         Square Root
7             PctBlack                 log

We note that fixing skewness might not resolve the issue of heteroscedasiticity but this can be done using "spreadLevelPlot" function.

# Statistical methodology - skewness

Key assumption in fitting a model is linearity. If predictor variables are skewed, this condition does not hold. To clean the data, it is good practice to remove any skewness from the data. 

A good visualisation is the use of histograms and density plots. This shows us which predictors are heavily skewed, which are acceptable and which are already symmetrical. We can use the "skewness" function from the "moments" package to quantify the skewness of each predictor variable. 

```{r}
#From Alex, James, and Remus 
#Checking if skewness falls in acceptable range using "skewness" function

cancer2$incidenceRate <- log(cancer2$incidenceRate)
hist(cancer2$incidenceRate, main = "Histogram of Transformed Incidence Rate")
plot(density(cancer2$incidenceRate), main= "Density estimate of Transformed Incidence Rate")
skewness(cancer2$incidenceRate) 

cancer2$PercentMarried<- cancer2$PercentMarried^(2)
hist(cancer2$PercentMarried, breaks = 30, main = "Histogram of transformed Percent Married")
plot(density(cancer2$PercentMarried), main="Density estimate of transformed Percent Married")
skewness(cancer2$PercentMarried)

cancer2$medIncomeFix <- log(cancer2$medIncome)
ggplot(cancer2, aes(medIncomeFix))+geom_density()
skewness(cancer2$medIncome)

cancer2$povertyPercentFix<- (cancer2$povertyPercent)^(1/3)
ggplot(cancer2, aes(cancer2$povertyPercentFix))+geom_density()
skewness(cancer2$povertyPercentFix)

cancer2$Avgfix <- log(cancer2$AvgHouseholdSize)
ggplot(cancer2, aes(Avgfix)) + geom_density()
skewness(cancer2$Avgfix)

cancer2$UnempFix <- sqrt(cancer2$PctUnemployed16_Over)
ggplot(cancer2, aes(UnempFix)) + geom_density()
skewness(cancer2$UnempFix)

cancer2$blackFix <- log(cancer2$PctBlack)
ggplot(cancer2, aes(blackFix)) + geom_density()
skewness(cancer2$blackFix)

```

Most predictors are positively skewed and most of the log and square root transformations fixed the skewness. However, we still observe that incidence rate has a high skewness that falls out of the acceptable range. This may be due to extreme values that differ from the median and that have high influence on the skewness. A box plot can be used to show this. 

```{r}
ggplot(cancer2, aes(cancer2$incidenceRate)) + geom_boxplot()
```

However, removing those values from the dataset might not be a good practice as further investigation might be needed as to whether or notthose values are due to some error. However, it might be beneficial to keep them as they reflect how incidence rates differ in some counties compared to the average incidence rate. This may be due to factors like demographic trend for example the proportion of aging population in the different counties therefore being at a higher risk of being diagnosed with cancer. 

We also observe that the incidence rate has a relatively higher correlation of 0.45 with deathrate (response variable) compared to other predictor variables. 

#Reference: 

https://www.cdc.gov/pcd/issues/2021/21_0006.htm (centre for disease control and prevention) 










