---
title: "ST404 A1 Remus 1st"
author: "Remus"
date: "2022/2/4"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## ST404

We are looking at 4 variables: povertyPercent, PctPrivateCoverage, PctEmpCoverage, PctPublicCoverage.

## Loading the data

```{r loading data and packages, echo = FALSE}
load("cancer.Rdata")
library(dplyr)
library(ggplot2)
library(tidyr)
library(GGally)
library(car)
df<-cancer
DF<-cancer[,c(5, 12:14)]
```

## Summary of the data
```{r summary, echo=FALSE}
summary(DF)
```

## Missing data
```{r missing data}
print("The sum of missing values in these 4 variables are:")
sum(is.na(DF))
```


## Univariate Plots
```{r univariate, echo = FALSE}
DF %>% gather() %>%
  ggplot(aes(x = value)) + facet_wrap(~key, scales = "free") +
  geom_histogram(col = c("blue"), fill = "blue", alpha=0.6)
```
We can see that the three Coverage variables are symmetrical but povertyPercent is right-skewed.
A transformation is needed to tackle this skewness.

## Predictors and Response
```{r predictor and response}
ggpairs(df[,c(5,12:14,18)], mapping  = aes(alpha = 0.3))  
```
The three coverage variables are highly correlated as expected. However, the
deathRate is not significantly correlated to any of the coverage variables.
The correlation coefficicent between deathRate and the coverage variables do not
exceed 0.45.

## Transformation
Try cube rooting.
```{r fix skew}
DF$povertyFIX <- (DF$povertyPercent)^(1/3)
ggplot(DF, aes(povertyFIX))+geom_density()
```



## Outliers
First we use box-plots to see if there are any suspicous value.

```{r boxplot, echo=FALSE}
DF %>% gather() %>%
  ggplot(aes(x = value)) + facet_wrap(~key, scales = "free") +
  geom_boxplot(shape=1, alpha=0.3)
```
From the box-plots, all four variables have a quite a big number of extreme values.
We need further investigations.


```{r iqr, echo=FALSE}
IQRpov<- quantile(DF$povertyFIX, 0.75) - quantile(DF$povertyFIX, 0.25)
UPpov<- quantile(DF$povertyFIX, 0.5) + 1.5*IQRpov
LOWpov<- quantile(DF$povertyFIX, 0.5) - 1.5*IQRpov

IQRprivate<- quantile(DF$PctPrivateCoverage, 0.75) - quantile(DF$PctPrivateCoverage,0.25)
UPprivate<- quantile(DF$PctPrivateCoverage, 0.5) +1.5*IQRprivate
LOWprivate<- quantile(DF$PctPrivateCoverage, 0.5) -1.5*IQRprivate

IQRemp<- quantile(DF$PctEmpPrivCoverage, 0.75) - quantile(DF$PctEmpPrivCoverage,0.25)
UPemp<-  quantile(DF$PctEmpPrivCoverage, 0.5) +1.5*IQRemp
LOWemp<- quantile(DF$PctEmpPrivCoverage, 0.5) -1.5*IQRemp

IQRpub<- quantile(DF$PctPublicCoverage, 0.75) - quantile(DF$PctPublicCoverage,0.25)
UPpub<-  quantile(DF$PctPublicCoverage, 0.5) +1.5*IQRpub
LOWpub<- quantile(DF$PctPublicCoverage, 0.5) -1.5*IQRpub

DF_IQRoutlier<- subset(DF, (povertyFIX<= LOWpov | povertyFIX >= UPpov) |
                           (PctPrivateCoverage<= LOWprivate | PctPrivateCoverage >= UPprivate)|
                           (PctEmpPrivCoverage<= LOWemp | PctEmpPrivCoverage >= UPemp)|
                           (PctPublicCoverage<= LOWpub | PctPublicCoverage >= UPpub))
```


## Hetroscedascity
We can look at the residuals plots of these variables in simple models.

```{r hetroscedascity, echo=FALSE}
par(mfrow=c(2,2))
plot(lm(deathRate~povertyPercent, data = cancer), 1)
title(sub="PovertyPercent")
plot(lm(deathRate~PctPrivateCoverage, data = cancer), 1)
title(sub="PctPrivateCoverage")
plot(lm(deathRate~PctEmpPrivCoverage, data = cancer), 1)
title(sub="PctEmpPrivCoverage")
plot(lm(deathRate~PctPublicCoverage, data = cancer), 1)
title(sub="PctPublicCoverage")
```
There is potential heteroscedascity in PctEmpPrivCoverage.
```{r ncv}
ncvTest(lm(deathRate~PctEmpPrivCoverage, data = cancer))
spreadLevelPlot(lm(deathRate~PctEmpPrivCoverage, data = df))
```
Try power=5
```{r hetroscedascity fix}
plot(lm(deathRate^(5)~PctEmpPrivCoverage, data = df),1)
ncvTest(lm(deathRate^5~PctEmpPrivCoverage, data = df))
```
```{r Multicollinearity}
vif(lm(deathRate~PctPrivateCoverage+PctEmpPrivCoverage+ PctPublicCoverage, data = df))
vif(lm(deathRate~PctPrivateCoverage+PctEmpPrivCoverage+ PctPublicCoverage+ povertyPercent, data = df))
```
Although VIF are not very big,the three healthcare coverage variables are
clearly correlated.
I would discard PctPrivateCoverage and PctEmpPrivCoverage in fitting a model.
I leave PctPublicCoverage because it is the least correlated variable with
povertyPercent among the healthcare coverage variables.
