---
title: "ST404 Assignment 1 Alex"
output: pdf_document
urlcolor: blue
fontsize: 11pt
geometry: margin = 1in
---

```{r setup, include = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
load("cancer.rdata")
library(VIM)
library(GGally)
library(usmap)
library(car)
library(stringr)
library(knitr)
```

# ST404 Assignment 1 Alex

## Checking the summary and initial EDA

\scriptsize

```{r Summary, echo = FALSE}
summary(cancer)
```

\normalsize

There are some missing values in PctEmployed16_Over which need to be checked.

The minimum value in AvgHouseholdSize is very small which is suspicious and should be immediately investigated.

```{r Sus Plot, echo = FALSE, fig.cap = "\\label{fig:figs}deathRate vs AvgHouseholdSize", fig.width=5, fig.asp=0.68, fig.align="center"}
plot(deathRate ~ AvgHouseholdSize, data = cancer, main = "deathRate against AvgHouseholdSize")
```

From the above plot we note that there are many extremely suspicious points with small AvgHouseholdSize.

We identify one of these points and investigate it:

```{r Extracting suspicious point, echo = FALSE}
kable(subset(cancer, AvgHouseholdSize < 0.5)[1,c("Geography", "AvgHouseholdSize")])
```

To check the validity of this data point we find an alternate source of the data at:

https://data.census.gov/cedsci/table?q=average%20household%20size&g=0500000US54003&y=2013&tid=ACSST1Y2013.S1101

We note that this data recording AvgHouseholdSize in the same year as our data lists the size at 2.61. This is completely different and this is similar for other small values in our dataset.

Hence, these are very likely incorrectly inputted data points and as there is only a small proportion of them we should treat them as missing data and then test to see whether they are MCAR.

```{r Replace small with NA}
cancer1 <- cancer
cancer1$AvgHouseholdSize[which(cancer1$AvgHouseholdSize < 0.5)] <- NA
```

## Missing values check

Now that we have replaced the small values with NAs we can test the data to see what kind of missing values we have.

```{r Miss, echo = FALSE, fig.cap = "\\label{fig:figs}Pbox plots showing difference between missing and non-missing data", results = "asis", fig.align = "center", fig.height = 3, fig.width = 7}
par(mfrow = c(1,4))
for (i in c(2:4, 18)){ 
  pbox(cancer1,pos=i) 
}
```

We use the pbox() function from the VIM package to check what these missing values represent.

From the above plots we note that the box plots with the missing data do not look significantly different from those without. The pbox plots for the other variables look similar to this so we conclude that the data that is missing is MCAR.

With our data we could replace all the data with an alternate source but as the proportion of missing data points is so small and it is MCAR it is safe to just remove the rows with missing data from our data set.

```{r remove miss}
cancer2 <- na.omit(cancer1)
```

## My allocation

```{r Plots against deathRate, fig.cap = "\\label{fig:figs}Plots showing deathRate against other variables", fig.width=5, fig.asp=1, fig.align="center", echo = FALSE}
par(mfrow = c(2,2))
with(cancer2, scatter.smooth(deathRate~AvgHouseholdSize, col = "red"))
with(cancer2, scatter.smooth(deathRate~PctEmployed16_Over, col = "red"))
with(cancer2, scatter.smooth(deathRate~PctUnemployed16_Over, col = "red"))
with(cancer2, scatter.smooth(deathRate~PctBlack, col = "red"))
```

From the bivariate plots there is definite heteroskedasticity in pctBlack and for AvgHouseholdSize we see some non linearity. We see a concave shape so advising a more compex model, perhaps with a quadratic term might be advisable as the data is not monotonic.


```{r historgrams for skew, echo = FALSE, fig.cap = "\\label{fig:figs}Histograms of our predictor variables", fig.width=5, fig.asp=0.68, fig.align="center"}
par(mfrow = c(2,3))
with(cancer2, hist(AvgHouseholdSize))
with(cancer2, hist(PctEmployed16_Over))
with(cancer2, hist(PctUnemployed16_Over))
with(cancer2, hist(PctBlack))
with(cancer2, hist(deathRate))
```

```{r BoxPlots, echo = FALSE, fig.cap = "\\label{fig:figs}BoxPlots of our variables", fig.align="center"}
par(mfrow = c(2,3))
with(cancer2, boxplot(AvgHouseholdSize, main = "AvgHouseholdSize"))
with(cancer2, boxplot(PctEmployed16_Over, main = "PctEmployed16_Over"))
with(cancer2, boxplot(PctUnemployed16_Over, main = "PctUnemployed16_Over"))
with(cancer2, boxplot(PctBlack, main = "PctBlack"))
with(cancer2, boxplot(deathRate, main = "deathRate"))
```

## Analysis of the above plots

### Scatter Plots

From the bivariate plots there is definite heteroskedasticity in pctBlack and for AvgHouseholdSize we see some non linearity. We see a concave shape so advising a more complex model, perhaps with a quadratic term might be advisable as the data is not monotonic.

For heteroskedasticity we would need to perform further tests after fititng a model to check what kind of transformation we'd need to fix it.

From the scatter plots there are no clear outliers, we'd need either some box plots or to look at cook's distance to identify that.


### Histograms

Massive right skew for pctBlack. PctUnemployed and AvgHouseholdSize are also a little right skew.
I Recommend a log transform for pctBlack and sqrt transforms for pct unemployed and avg household size.

```{r Transforms, fig.cap = "\\label{fig:figs}Our transformed histograms", fig.width=5, fig.asp=0.68, fig.align="center"}
par(mfrow = c(1,3))
with(cancer2, hist(sqrt(AvgHouseholdSize), main = "Transformed AvgHouseholdSize"))
with(cancer2, hist(sqrt(PctUnemployed16_Over), main = "Transformed PctUnemployed16_Over"))
with(cancer2, hist(log(cancer2$PctBlack), main = "Transformed PctBlack"))
```

### Box Plots

Our Box Plots show we have quite a number of what we would consider outliers accross all our variables. This does not necessarily mean that they should be removed as we do not know their influence yet due to not fitting a model.

We have a severe amount of outliers in PctBlack according to our box plot. This could be due to the very long tail as shown in the scatter plot above.

### BIG MAP

```{r map, eval = FALSE, echo = FALSE, fig.cap = "\\label{fig:figs}A map showing deathRate accross the United States", fig.width=5, fig.asp=0.68, fig.align="center"}
cancer4 <- cancer2
uwu2 <- str_split(cancer4$Geography, pattern = ", ")
cancer4$state <- rep(0, length(cancer4$Geography))
for(i in 1:length(cancer4$Geography)){
  cancer4$state[i] <- uwu2[[i]][2]
}

cancer4$county <- rep(0, length(cancer4$Geography))
for(i in 1:length(cancer4$Geography)){
  cancer4$county[i] <- uwu2[[i]][1]
}
cancer4$county[159] <- "Dona Ana County"
cancer4$county[775] <- "La Salle Parish"
cancer4$fips <- rep("0", length(cancer4$Geography))
for(i in 1:length(cancer4$Geography)){
  cancer4$fips[i] <- fips(cancer4$state[i], cancer4$county[i])
}
deathMap <- subset(cancer4, select = c("fips", "county", "deathRate"))
plot_usmap(data = deathMap,regions = "counties", values = "deathRate", include = cancer4$fips, color = "red") +
  scale_fill_continuous(low = "yellow", high = "red", name = "Death Rate", label = scales::comma) +
  labs(title = "Death rates in the United States") +
  theme(legend.position = "right")
```

From the map we note that the deathRate appears to be higher in the mid-eastern United States