---
title: "variables no lm"
author: "Remus"
date: "2022/2/8"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We look at povertyPercent, PctPrivateCoverage, PctEmpPrivCoverage, PctPublickCoverage

## Loading data and packages

```{r Loading, message = FALSE}
load("cancer.RData")
load("cancer2.RData")
library(dplyr)
library(GGally) 
library(ggplot2)
library(tidyr)  
library(car)
```

## Using Alex's cancer2 (removed missing or incorrect values)

```{r cancer2}
cancer1 <- cancer
susPoint <- which(cancer1$AvgHouseholdSize < 0.5)
cancer1$AvgHouseholdSize[susPoint] <- NA
cancer2 <- na.omit(cancer1)
```

## Scatter Plots
```{r Linearity, echo=FALSE}
cancer2 %>% gather(-deathRate,-Geography,-binnedInc, key="predictor", value="value") %>%
  ggplot(aes(x = value, y=deathRate)) + facet_wrap(~predictor, scales = "free") +
  geom_point(size=0.4) +
  ggtitle("Scatter plots of deathRate against predictor variables")
```


```{r Scatter smooth,echo=FALSE}
par(mfrow=c(2,2))
scatter.smooth(cancer2$povertyPercent, cancer2$deathRate, lpars=list(col = "red", lwd = 2), ylab="deathRate",xlab="povertyPercent")
scatter.smooth(cancer2$PctPrivateCoverage, cancer2$deathRate, lpars=list(col = "red", lwd = 2),
ylab="deathRate", xlab="PctPrivateCoverage")
scatter.smooth(cancer2$PctPublicCoverage, cancer2$deathRate, lpars=list(col = "red", lwd = 2),
ylab="deathRate",xlab="PctPublicCoverage")
scatter.smooth(cancer2$PctEmpPrivCoverage, cancer2$deathRate, lpars=list(col = "red", lwd = 2),
ylab="deathRate",xlab="PctEmpPrivCoverage")
```

We can see all four variables have fairlty straight fitted lines. This support
them being linear.
The data points of PctEmpPrivCoverage are getting closer to the fitted line
which shows that the variance is decreasing.
We might use spreadLevelPlot() to find appropriate power transformation to fix
this problem.


##Skewness 
```{r Skew ,echo=FALSE}
cancer3<- cancer2[,c(5,12:14)]
cancer3 %>% gather() %>%
  ggplot(aes(x = value)) + facet_wrap(~key, scales = "free") +
  geom_histogram(col = c("blue"), fill = "blue", alpha=0.6)
```
From the histograms, PctPrivateCoverage is slightly left skewed and povertyPercent
is right skewed.
Transformations are needed.
Try cubeth root for povertyPercent.

```{r Transform1, echo=FALSE}
ggplot(cancer2, aes(x=povertyPercent^(1/3)))+
  geom_histogram(col = c("blue"), fill = "blue", alpha=0.6)+
  ggtitle("Histogram of cubeth root of povertyPercent")+
  ylab("Frequency")
```
The skewness in povertyPercent is removed in the cubeth rooting transfomation.
Next try squaring for PctPrivateCoverage.

```{r Trasnform2, echo=FALSE}
ggplot(cancer2, aes(x=PctPrivateCoverage^(2)))+
  geom_histogram(col = c("blue"), fill = "blue", alpha=0.6)+
  ggtitle("Histogram of square of PctPrivateCoverage")+
  ylab("Frequency")

```
The skewness is in PctPrivateCoverage is removed by the square transformation.


##Outliers

```{r boxplots, echo=FALSE}
cancer3 %>% gather() %>%
  ggplot(aes(x = value)) + facet_wrap(~key, scales = "free") +
  geom_boxplot(shape=1, alpha=0.5, col="blue")+
  coord_flip()
```
All four variables have points lying outside the boxes.
Note that PctPrivateCoverage have points lying below the box and
povertyPercent have points lying above the box only.
This shows potential outliers in particular in PctPrivateCoverage and povertyPercent.
Further discoveries and decisions on the outliers should be done when fitting a 
model.

## Multicollinearity

Poverty with PctUnemployed_Over16, PctEmployed_Over16, PctPrivateCoverage, PctEmpPrivCoverage, PctPublicCoverage
```{r cor, echo=FALSE}
round(cor(cancer2[,c(5,10:14)])[1,], digits = 2)
```
Correlation is at least 0.65. 
In particular -0.74 with PctEmployed16_Over, -0.82 with PctPrivateCoverage.

Proceed using Pearson Correlation Coefficent test.
We get p-values 2.2e-16 for all five tests.
We can conclude povertyPercent is significant correlated to these five variables.

```{r cor test,echo=FALSE}
cor.test(cancer2$povertyPercent, cancer2$PctEmployed16_Over)
cor.test(cancer2$povertyPercent, cancer2$PctUnemployed16_Over)
cor.test(cancer2$povertyPercent, cancer2$PctPrivateCoverage)
cor.test(cancer2$povertyPercent, cancer2$PctEmpPrivCoverage)
cor.test(cancer2$povertyPercent, cancer2$PctPublicCoverage)
```



