---
title: "Missing Values Alex"
author: "Alex"
date: "08/02/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
load("cancer.rdata")
summary(cancer)
```

There are some missing values in PctEmployed16_Over which need to be checked.
Before that is checked though we should note that the outliers in AvgHouseholdSize should be treated as missing values as they are wrongly inputted (reference).

```{r}
cancer1 <- cancer
cancer1$AvgHouseholdSize[cancer1$AvgHouseholdSize < 0.5] <- NA
```

Now I use the VIM package and the pbox() function to show that the missing data are all MCAR and can thus be easily dealt with.

```{r, message = FALSE}
library(VIM)
par(mfrow = c(1,4))
for(i in 2:5){
  pbox(cancer1, pos = i)
}
```

```{r}
par(mfrow = c(1,4))
for(i in 6:9){
  pbox(cancer1, pos = i)
}
```


```{r}
par(mfrow = c(1,4))
for(i in 10:13){
  pbox(cancer1, pos = i)
}
```

```{r}
par(mfrow = c(1,5))
for(i in 14:18){
  pbox(cancer1, pos = i)
}
```

From the pbox plots we have evidence that the missing values interrogated are MCAR. Since the proportion of rows with missing values is small I recommend simply deleting the rows with missing values as it should not affect the validity of our analysis on the whole dataset as the smaller sample is still representative.

```{r}
cancer2 <- na.omit(cancer1)
```

Our other option is to use the reference to fill in all our other missing values. Despite this being possible it is also un-necessary as the data is MCAR and won't significantly affect our analysis.



