---
title: "ST404 Small Report"
author: "Alex"
date: "03/02/2022"
output: 
  pdf_document:
    fig_caption: yes
---

# ST404 Project

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First we load the dataset and required libraries

```{r packages, warning = FALSE}
load("cancer.rdata")
library(VIM)
library(GGally)
library(usmap)
library(car)
library(stringr)
```

## Summary

First lets have a look at a summary of all the data:

```{r summary}
summary(cancer)
```

We notice that there is one categorical variable (binnedInc) and one ID variable (Geography). Geography shouldn't be used as an explanatory variable but could be used as a visual aid.

We also note that PctEmployed16_Over has 152 missing values that should be investigated.

## Univariate Plots:

```{r uni, echo = FALSE, warning = FALSE}
par(mfrow = c(2,2))
hist(cancer$AvgHouseholdSize)
hist(cancer$PctEmployed16_Over)
hist(cancer$PctBlack)
hist(cancer$PctUnemployed16_Over)
```

We note that PctBlack has very large right skew and will need transforming to fix this.

We also note that there are some very suspicious low values in AvgHouseholdSize which are outliers. We create a new dataset containing these suspicious values.

```{r strange}
sus <- subset(cancer, AvgHouseholdSize < 0.5)
head(sus)
```

I will use an alternate source to check if the AvgHousehold Size is correct for Berkeley County.

```{r Berk}
sus[1,"AvgHouseholdSize"]
```

Reference:
https://data.census.gov/cedsci/table?q=average%20household%20size&g=0500000US54003&y=2013&tid=ACSST1Y2013.S1101

From the same census data we have that the average household size of this county is 2.61. This indicates the data has been incorrectly inputted into this dataset or the data is missing. Either way, this incorrect data could cause the distribution to be unstable and should be considered as missing.

```{r NA implement}
cancer1 <- cancer
susPoint <- which(cancer1$AvgHouseholdSize < 0.5)
cancer1$AvgHouseholdSize[susPoint] <- NA
summary(cancer1)
```


## Missing Values and Strange Values:

There are now two predictor variables with what we would consider missing or strange values. We have to decide what kind of missing values these are. We can use plots from the VIM package to do this.

```{r Miss}
par(mfrow = c(1,4))
for (i in c(2:5)){ 
  pbox(cancer1,pos=i) 
}
```

From the plots the difference in boxplots between missing values and non missing is quite similar across the board. We can consider these missing values MCAR and should just omit those rows with missing values as there isn't that many so should not cause our data to be unstable.

```{r NA Omit}
cancer2 <- na.omit(cancer1)
```

## Comparing Predictors and Response

```{r}
ggpairs(cancer2, columns = c(8,10,11,15,18), mapping = aes(alpha = 0.6),
        upper = list(continuous=wrap("cor", size=2.5, digits=2, stars=FALSE)),
        lower = list(continuous=wrap("points", size = 0.5)),
        diag = list(continuous="densityDiag"),
        progress = FALSE) +
  theme(text=element_text(size=6))
```

Unemployed and Employed 16 or over have high correlation with each other.


Below we have some plots of our variables against death rate:

```{r Plot Death, echo = FALSE,fig.cap = "\\label{fig:figs}Predictors vs Death Plots"}
par(mfrow = c(2,2))
plot(deathRate~AvgHouseholdSize, data = cancer2)
plot(deathRate~PctEmployed16_Over, data = cancer2)
plot(deathRate~PctUnemployed16_Over, data = cancer2)
plot(deathRate~PctBlack, data = cancer2)
```

PctBlack is incredibly right skewed, we can try a log transformation to fix it. 

```{r Fixing Black}
cancer2$blackFix <- log(cancer2$PctBlack)
ggplot(cancer2, aes(blackFix)) + geom_density()
```

PctUnemployed16_Over is also slightly right skewed. We can try a square root transformation to fix this.

```{r Fix Unemployment}
cancer2$UnempFix <- sqrt(cancer2$PctUnemployed16_Over)
ggplot(cancer2, aes(UnempFix)) + geom_density()
```

Average household size maybe needs fixing

```{r Size Fix}
cancer2$Avgfix <- log(cancer2$AvgHouseholdSize)
ggplot(cancer2, aes(Avgfix)) + geom_density()
```

We have evidence of multicolinearity between Unemployment and Employment.


```{r Residual Plots, echo = FALSE, fig.cap = "\\label{fig:figs}Residual Plots"}
par(mfrow = c(2,2))
plot(lm(deathRate~AvgHouseholdSize, data = cancer2), 1)
plot(lm(deathRate~PctUnemployed16_Over, data = cancer2), 1)
plot(lm(deathRate~PctEmployed16_Over, data = cancer2), 1)
plot(lm(deathRate~PctBlack, data = cancer2), 1)
boxCox(lm(deathRate~AvgHouseholdSize, data = cancer2))
```

We have evidence of a lot of heteroscedasticity for PctBlack, we can use a spread level plot for a suggested transformation

```{r Spread Level}
spreadLevelPlot(lm(deathRate~PctBlack, data = cancer2))
ncvTest(lm(deathRate~PctBlack, data = cancer2))
```

We try cubing the response to fix the heteroscedasticity

```{r Cube}
plot(lm(deathRate^3~PctBlack, data = cancer2), 1)
spreadLevelPlot(lm(deathRate^3~PctBlack, data = cancer2))
```

This seems to have mostly fixed heteroscedasticity

This is a different transformation than the one used to fix the density though

There is also some heteroscedasticity present for employed 16 or over.

```{r More Spread}
ncvTest(lm(deathRate~PctEmployed16_Over, data = cancer2))
spreadLevelPlot(lm(deathRate~PctEmployed16_Over, data = cancer2))
```

```{r Fixing}
plot(lm(deathRate^(-2)~PctEmployed16_Over, data = cancer2), 1)
spreadLevelPlot(lm(deathRate^(-2)~PctEmployed16_Over, data = cancer2))
```

## Multicolinearity

We have high correlation between employed 16 and over and unemployed 16 and over. We will test to see if both are necessary.

```{r Multi}
vif(lm(deathRate~PctEmployed16_Over + PctUnemployed16_Over, data = cancer2))
```

Variance inflation factors are low though, so we could potentially use both as they aren't measuring the exact same thing.

```{r}
cancer3 <- cancer
uwu2 <- str_split(cancer3$Geography, pattern = ", ")
cancer3$state <- rep(0, length(cancer3$Geography))
for(i in 1:length(cancer3$Geography)){
  cancer3$state[i] <- uwu2[[i]][2]
}

cancer3$county <- rep(0, length(cancer3$Geography))
for(i in 1:length(cancer3$Geography)){
  cancer3$county[i] <- uwu2[[i]][1]
}
cancer3$county[167] <- "Dona Ana County"
cancer3$county[821] <- "La Salle Parish"
cancer3$fips <- rep("0", length(cancer3$Geography))
for(i in 1:length(cancer3$Geography)){
  cancer3$fips[i] <- fips(cancer3$state[i], cancer3$county[i])
}
deathMap <- subset(cancer3, select = c("fips", "county", "deathRate"))
plot_usmap(data = deathMap,regions = "counties", values = "deathRate", include = cancer3$fips, color = "red") +
  scale_fill_continuous(low = "yellow", high = "red", name = "Death Rate", label = scales::comma) +
  labs(title = "Death rates in the states") +
  theme(legend.position = "right")
```

The map above gives a nice illustration of what death rate looks like accross the United States, whites indicate that county was missing from the original dataset.

My recommendations for fixing would be apply the above transforms to fix skew, also remove the suspect and missing data as well to make sure our data isn't unstable.

```{r Binned Check}
#vif(lm(deathRate~binnedInc + medIncome, data = cancer2))
ols_coll_diag(lm(deathRate~binnedInc + medIncome, data = cancer2))
```

```{r}
par(mfrow = c(2,2))
plot(lm(deathRate~medIncome, data = cancer2))
```

```{r}
plot(lm(deathRate~AvgHouseholdSize, data = cancer2), 1)
```
```{r}
vif(lm(deathRate ~ PctEmpPrivCoverage + PctPrivateCoverage, data = cancer2))
```

```{r}
plot(lm(deathRate ~ incidenceRate, data = cancer2), 4)
cancer2[267,]
```

```{r}
vif(lm(deathRate ~ PercentMarried + PctMarriedHouseholds, data = cancer2))
```


```{r}
plot(PercentMarried~PctMarriedHouseholds, data = cancer2)
```


```{r}
boxplot(cancer2$binnedInc)
#lines(cancer2$medIncome, cancer2$binnedInc)
```

```{r}
plot(medIncome~binnedInc, data = cancer2)
```

```{r}
marginalModelPlots(lm(deathRate~AvgHouseholdSize, data = cancer2))
```

```{r}
boop <- lm(sqrt(deathRate)~AvgHouseholdSize, data = cancer2)
plot(boop, 1)
```


```{r}
boxCox(boop)
```

```{r missing visualisation}
aggr(cancer1, prop = FALSE, combined = TRUE, numbers = TRUE, 
                sortVars = TRUE, sortCombs = TRUE)
```




